Revealing Benefits, Costs, and Tradeoffs of Spatial Management in Marine Systems Arrising from Coupled Economic and Ecological Dynamics

### 2023-01-05 03_marlin_paper.qmd
echelle edits

`marlin` also allows users to specify closed fishing seasons for one or more
species in the system. During closed seasons, no effort is allocated to
patches that would result in capture of the covered species, potentially
resulting in zero effort if the species cannot be avoided. Under these
conditions, the amount of effort in the next open season is a function
of the effort and profits in the last open fishing season. Lastly,
managers can specify locations of no-take MPAs, which could in theory
change in size and location. When MPAs are implemented, fishing effort
that used to operate inside the MPAs can either leave the fishery, or be
redistributed outside the MPA (the default behavior).

@@

Each fishing fleet (*f*) generates catches or yield (*Y*), revenues
(*R*), costs (*C*) from individual species (*s*) that it targets in a
given time step (*t*) and patch (*p*). The fleet then makes decisions
around how much and where to fish, subject to regulatory constraints,
based on the total profitability across all species in space and time.

Revenues for each fleet in a time step are a function of the total
amount caught of each species and its price ($\Pi$). The amount
caught is a function of the selectivity at age of each species for each
fleet ($\alpha$), the fishing efficiency of the fleet for that that
species (also called "catchability", *q*), the amount of fishing effort
of the fleet in question in that patch in a given time step (*E*), and
the total fishing mortality at age (including all other fleets) for that
particular species in that patch and time (*u*).

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of *advection* (follow currents), and *taxis* (active movement towards preferred habitat) and *diffusion* (essentially remaining variation in movement not explained by diffusion or taxis). For now, `marlin` focuses just on the
diffusion and taxis components of this model, though future extensions
could easily incorporate advection vectors from oceanographic models. In
this way, `marlin` allows users to run anything from a simple Gaussian
dispersal kernel up to a system governed by species that passively
diffuse around a dynamic thermal range.

@@

The individual components of the diffusion matrix are filled based on an adjacency matrix, which defines whether two cells are both adjacent and water (as opposed to land or another physical barrier), and a diffusion rate $D$ definend in units of length of the side of a cell per unit of time. For example, if we are defining the time units as years and the distance units as kilometers, for a tuna $D$ might be 1,000 $\frac{KM}{Year}$. The $\Delta_{t}$ and $\Delta_{p}$ parameters then translate the movement rate $D$ to match the time step and cell size used in a simulation. This "scale free" parameterization means that appropriate value of $D$ can be identified for a species and then set, regardless of the time step or cell size used in the simulation model itself.

@@

where *BH* is the Beverton-Holt recruitment function, *ssb* is
spawning-stock-biomass. $l_a$ is the mean length at at age, $l_{sel}$ is
the length at which on average 50% of individuals are selected by the
fishery, and $\delta_{sel}$ are the additional units of length at which
on average 95% of fish are selected by the fishery.

@@

Fisheries management is often concerned with measuring stock status
relative to maximum sustainable yield (MSY) based reference points,
though the exact level of stock status desired by management or societal
objectives may vary widely. MSY based reference points present a problem
for a multi-fleet and spatial-temporal model such as `marlin`. MSY and
the fishing mortality rate that would produce MSY, $F_{MSY}$ are a
function of fishery selectivity. Fishery selectivity in this model can
vary by fleet, and species can be distributed unevenly in both space and
time. This means that the net effective fishing selectivity on a species
can vary depending on the dynamics at a given moment, making the
definition of an equilibrium concept such as MSY challenging.

@@

2.  Target_fishing (*target_fishing*): MPAs locations are prioritized
    proportional to fishery catches. So, patches with high total
    catches are prioritized over patches with low catches.

To our knowledge, MPAs are never randomly placed; their locations are
decided based on stakeholder negotiation given available information and
objectives. In theory, the design of MPA networks can be optimized
through the use of a modeling framework, and indeed conditional on the
validity of the model itself this process is likely to produce better
outcomes than manually-designed strategies [@rassweiler2012;
@rassweiler2014]. However, designing optimal MPA networks becomes
increasingly difficult as the range of objectives and the complexity of
the model increase. Therefore, while `marlin` provides some support for
[optimization](https://danovando.github.io/marlin/articles/optimize-mpas.html),
we focus here on the design and performance of these more "rule of
thumb" design strategies that may be more accessible to a wider range of
users. We allow all MPAs to be designed in a "mosaic" fashion, but users
can easily extend the analysis to for example compare outcomes between
"contiguous" and "mosaic" MPA designs [@pons2022].

@@

In our coral reef example, we model the dynamics of four species: a
grouper (*Epinephelus fuscoguttatus*), a shallow-reef snapper (*Lutjanus
malabaricus*), a deep-water snapper (*Pristipomoides filamentosus*), and
a reef shark (*Carcharhinus amblyrhynchos*). The simulated groupers
undergo a mass migration to a spawning aggregation once per year,
followed by the sharks. Shallow-reef snappers stay in reefs closer to
shower year-round, while deep-water snappers stay in the deeper reefs
(@fig-coral).

These species are targeted by two different fleets. Fleet one primarily
targets the grouper and nearshore snapper populations, but will land any
incidentally captured sharks. Fleet one has a dome shaped selectivity
pattern for both grouper and shallow-reef snapper due to demand for
"plate sized" fish [@kindsvater2017], and logistic selectivity for
sharks, as any shark caught is retained. This fleet is a subsistence
fleet, where the local community takes every possible opportunity to
fish that they can, and as such we model it as a "constant effort"
fishery.

The second fleet primarily targets the deep-water snapper population.
While plate-sized snapper are still prized from this population, we
model fleet two's selectivity as logistic due to high levels of discard
mortality for deep-water snapper resulting from barotrauma. Fleet two
catches some grouper, though less than fleet one, and receives no price
for sharks due to the requirements of a certification program through
which they sell their deep-water snapper, though accidental captures do
occur, which results in mortality. This fleet operates under open-access
dynamics, as fishing is not the primary means of subsistence and so
short-term effort expands and contracts in response to profitability of
the primarily grouper-driven fishery.

@@

We use `marlin` to simulate the outcomes for both food production and
conservation for each of the species and both of the fleets as a
function of both MPA size and MPA design strategy. For this exercise,
MPAs are placed with perfect information and have no design constraints
for continuity (i.e. the selected MPA design can be as patchwork as
dictated by the data). We ran the simulation on a quarterly level, meaning that $\Delta_t = 1/ 4$ and set the area of each cell to an arbitrary value of 1 (meaning $\Delta_p = 1$ ).

@@

We model our pelagic use case loosely on the characteristics of the
Western and Central Pacific Ocean tuna fisheries. We simulate
trajectories of `r n_distinct(names(blue_fauna))` species commonly
caught in the region, including both the highly abundant skipjack tuna
to the heavily depleted oceanic whitetip shark (*Carcharhinus
longimanus*) (@fig-blue-ssb). We use publicly available data on
catch-per-unit-effort of each of these species from the WCPO as a proxy
for baseline habitat distributions, noting that where possible
fishery-independent abundance indices would be preferable (@fig-blue).

@@

For this exercise, we focused on using `marlin` to assess resilience of
the selected "rule of thumb" MPA design strategies to a climate-driven
range shift. Specifically, we simulate an extreme example where the
centroid of each population shifts northward at a rate of 0.2 patches
per year, over a `r experiment_years` year time horizon (@fig-blue-ssb).
We designed MPA networks given the conditions in the starting year, and
then held that network constant over the years of the experiment,
running one simulation with and another without the climate-drive range
shift. We then compared the effects of this range shift on food
production and conservation outcomes from MPA networks designed based on
the pre-range shift world. We ran the simulation on a quarterly level, meaning that $\Delta_t = 1/4$ and set the area of each cell to an arbitrary value of 1 (meaning $\Delta_p = 1$ ).

@@

These macro-level changes in total yield and conservation outcomes mask
some internal tradeoffs. Nearly all of the food production benefits from
this MPA network accrued to fleet one, while catches by fleet two were generally
negatively impacted under both scenarios. For example, the MPA sizes and designs that result in the largest increase in yields for fleet one would produce a near total
loss of fish catch for fleet two.

@@

The role of this paper is not to conduct a review of the many
tools for spatial-temporal modeling available in the literature.
However, we highlight here the specific gaps that we feel `marlin` fills
in the modeling literature and in the policy support toolbox using some
selected publications. More end-to-end models such as ATLANTIS
[@audzijonyte2019] , OSMOSE [@shin2001], Ecopath with Ecosim
[@christensen2004], POSEIDON [@bailey2019], or SEAPODYM [@lehodey2008]
are capable of representing tremendous amounts of complexity, but can be
extremely time consuming to design and run.

Tools such as virtualspecies
[@leroy2016], STEPS [@visintin2020], RangeShifter [@bocedi2014], and
[SMS](https://www.aquaveo.com/software/sms-surface-water-modeling-system-introduction)
can be efficiently constructed to model the dispersal and distribution
of species in space and time as a function of environmental covariates.
However, these typically require specifying a covariate-response curve
without explicitly acknowledging how this arises from habitat-specific
movement and demography (virtualspecies), or model movement using a
dispersal kernel (RangeShifter, STEPS) or a least-cost path algorithm
(SMS, STEPS). By contrast, we specify a continuous-time Markov chain
movement model, which integrates multiple paths (including their
path-dependent probability based on intervening habitat types and
species preferences) while using scale-free parameters that can be
measured experimentally in laboratory or tagging studies.

Simplified bio-economic models such as those used in @hastings2017,
@sala2021, and @cabral2019 can be applied at scale and provide
analytically tractable results, but as a result must abstract over many
bio-economic dynamics that can be important for more tactical
applications. `marlin` complements the existing spatial marine modeling
literature by allowing scientists, decision makers, and other
stakeholders to efficiently and productively examine the impacts of
realistic bio-economic dynamics on academic and applied problems.

>>>>

`marlin` also allows users to specify closed fishing seasons for one or more
species in the system. When all species in a simulation share the same closed seasons, no fishing occurs in those seasons. When not all species share the same closed seasons, fishing still occurs in the closed seasons, but the fleet avoids catching the species for which fishing is closed in a given season, assuming no discard mortality. When closed seasons are in effect, the amount of effort in the next open season is a function of the effort and profits in the last open fishing season. Lastly,
managers can specify locations of no-take MPAs, which could in theory
change in size and location. When MPAs are implemented, fishing effort
that used to operate inside the MPAs can either leave the fishery, or be
redistributed outside the MPA (the default behavior).

@@

Each fishing fleet (*f*) generates catches or yield (*Y*), revenues
(*R*), and costs (*C*) from fishing individual species (*s*) that it targets in a
given time step (*t*) and patch (*p*). The fleet then makes decisions
around fishing locations and intensity, subject to regulatory constraints,
based on the total profitability across all species in space and time.

Revenues for each fleet in a time step are a function of the total
amount of each species caught and its price ($\Pi$). The amount
caught is a function of the selectivity at age of each species for each
fleet ($\alpha$), the fishing efficiency of the fleet for that that
species (also called "catchability", *q*), the amount of fishing effort
of the fleet in question in that patch in a given time step (*E*), and
the total fishing mortality at age (including all other fleets) for that
particular species in that patch and time (*u*).

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of *advection* (follow currents), and *taxis* (active movement towards preferred habitat) and *diffusion* (essentially remaining variation in movement not explained by advection or taxis). For now, `marlin` focuses just on the
diffusion and taxis components of this model, assuming that advection is zero, though future extensions
could easily incorporate advection vectors from oceanographic models. In
this way, `marlin` allows users to run anything from a simple Gaussian
dispersal kernel up to a system governed by species that passively
diffuse around a dynamic thermal range.

@@

The individual components of the diffusion matrix are filled based on an adjacency matrix, which defines whether two cells are both adjacent and water (as opposed to land or another physical barrier), and a diffusion rate $D$ defined in units of length of the side of a cell per unit of time. For example, if we are defining the time units as years and the distance units as kilometers, for a tuna $D$ might be 1,000 $\frac{KM}{Year}$. The $\Delta_{t}$ and $\Delta_{p}$ parameters then translate the movement rate $D$ to match the time step and cell size used in a simulation. This "scale free" parameterization means that appropriate value of $D$ can be identified for a species and then set, regardless of the time step or cell size used in the simulation model itself.

@@

where *BH* is the Beverton-Holt recruitment function, *ssb* is
spawning-stock-biomass, $l_a$ is the mean length at at age, $l_{sel}$ is
the length at which on average 50% of individuals are selected by the
fishery, and $\delta_{sel}$ are the additional units of length at which
on average 95% of fish are selected by the fishery.

@@

Fisheries management is often concerned with measuring stock status
relative to maximum sustainable yield (MSY) based reference points,
though the exact level of stock status desired by management or societal
objectives may vary widely. MSY based reference points present a problem
for a multi-fleet and spatial-temporal model such as `marlin`. MSY and
the fishing mortality rate that would produce MSY, $F_{MSY}$, are a
function of fishery selectivity. Fishery selectivity in this model can
vary by fleet, and species can be distributed unevenly in both space and
time. This means that the net effective fishing selectivity on a species
can vary depending on the dynamics at a given moment, making the
definition of an equilibrium concept such as MSY challenging.

@@

2.  Target Fishing (*target_fishing*): MPAs locations are prioritized
    proportional to fishery catches. Patches with high total
    catches are prioritized over patches with low catches.

To our knowledge, MPAs are never randomly placed; their locations are
decided based on stakeholder negotiation given available information and
objectives. In theory, the design of MPA networks can be optimized
through the use of a modeling framework, and depending on the
validity of the model, this process is likely to produce better
outcomes than manually-designed strategies [@rassweiler2012;
@rassweiler2014]. However, designing optimal MPA networks becomes
increasingly difficult as the range of objectives and the complexity of
the model increase. Therefore, while `marlin` provides some support for
[optimization](https://danovando.github.io/marlin/articles/optimize-mpas.html),
we focus here on the design and performance of these more "rule of
thumb" design strategies that may be more accessible to a wider range of
users. We allow all MPAs to be designed in a "mosaic" fashion, but users
can easily extend the analysis to compare outcomes between
"contiguous" (MPA is made up of one continuously connected block) and "mosaic" (MPAs can be separated in space) MPA designs [@pons2022].

@@

In our coral reef example, we model the dynamics of four species: a
grouper (*Epinephelus fuscoguttatus*), a shallow-reef snapper (*Lutjanus
malabaricus*), a deep-water snapper (*Pristipomoides filamentosus*), and
a reef shark (*Carcharhinus amblyrhynchos*). The simulated groupers
undergo a mass migration to a spawning aggregation once per year,
followed by the sharks. Shallow-reef snappers stay in reefs closer to
shore year-round, while deep-water snappers stay in the deeper reefs
(@fig-coral).

These species are targeted by two different fleets. Fleet one primarily
targets the grouper and nearshore snapper populations, but will land any
incidentally captured sharks. Fleet one has a dome shaped selectivity
pattern for both grouper and shallow-reef snapper due to demand for
"plate sized" fish [@kindsvater2017], and logistic selectivity for
sharks, as any shark caught is retained. This fleet is a subsistence
fleet, where the local community takes advantage of every possible opportunity to
fish, and as such we model it as a "constant effort"
fishery.

The second fleet primarily targets the deep-water snapper population.
While plate-sized snapper are still prized from this population, we
model fleet two's selectivity as logistic due to high levels of discard
mortality for deep-water snapper resulting from barotrauma. Fleet two
catches some grouper, though less than fleet one, and receives no price
for sharks due to the requirements of a certification program through
which they sell their deep-water snapper. Accidental captures (bycatch) do
occur, which results in mortality. This fleet operates under open-access
dynamics, as fishing is not the primary means of subsistence;
short-term effort expands and contracts in response to profitability of
the primarily grouper-driven fishery.

@@

We use `marlin` to simulate the outcomes for both food production and
conservation for each of the species and both of the fleets as a
function of both MPA size and MPA design strategy. For this exercise,
MPAs are placed with perfect information and have no design constraints
for continuity (i.e. the selected MPA design can be as patchwork as
dictated by the data). We ran the simulation on a quarterly level ($\Delta_t = 1/4$) and set the area of each cell to an arbitrary value of 1 ($\Delta_p = 1$).

@@

We model our pelagic use case loosely on the characteristics of the
Western and Central Pacific Ocean (WCPO) tuna fisheries. We simulate
trajectories of `r n_distinct(names(blue_fauna))` species commonly
caught in the region, including both the highly abundant skipjack tuna
to the heavily depleted oceanic whitetip shark (*Carcharhinus
longimanus*) (@fig-blue-ssb). We use publicly available data on
catch-per-unit-effort of each of these species from the WCPO as a proxy
for baseline habitat distributions, noting that where possible,
fishery-independent abundance indices would be preferable (@fig-blue).

@@

For this exercise, we focused on using `marlin` to assess resilience of
the selected "rule of thumb" MPA design strategies to a climate-driven
range shift. Specifically, we simulate an extreme example where the
centroid of each population shifts northward at a rate of 0.2 patches
per year, over a `r experiment_years` year time horizon (@fig-blue-ssb).
We designed MPA networks given the conditions in the starting year, and
then held that network constant over the years of the experiment,
running one simulation with and another without the climate-drive range
shift. We then compared the effects of this range shift on food
production and conservation outcomes from MPA networks designed based on
the pre-range shift world. We ran the simulation on a quarterly level ($\Delta_t = 1/4$) and set the area of each cell to an arbitrary value of 1 ($\Delta_p = 1$).

@@

These macro-level changes in total yield and conservation outcomes mask
some internal tradeoffs. Nearly all of the food production benefits from the simulated MPA networks accrued to fleet one, while catches by fleet two were generally
negatively impacted under both scenarios. For example, the MPA sizes and designs that result in the largest increase in yields for fleet one would produce a near total
loss of fish catch for fleet two.

@@

The role of this paper is not to conduct a review of the many
tools for spatial-temporal modeling available in the literature, each of which provide useful functionality for specific applications. However, we highlight here the specific gaps that we feel `marlin` fills
in the modeling literature and in the policy support toolbox using some
selected publications. More end-to-end models such as ATLANTIS
[@audzijonyte2019] , OSMOSE [@shin2001], Ecopath with Ecosim
[@christensen2004], POSEIDON [@bailey2019], or SEAPODYM [@lehodey2008]
are capable of representing tremendous amounts of complexity, but can be
extremely time consuming to design and run. `marlin` allows some of the realism of these end-to-end models while being simpler and faster to parameterize and run.

Tools such as virtualspecies
[@leroy2016], STEPS [@visintin2020], RangeShifter [@bocedi2014], and
[SMS](https://www.aquaveo.com/software/sms-surface-water-modeling-system-introduction)
can be efficiently constructed to model the dispersal and distribution
of species in space and time as a function of environmental covariates.
However, these typically require specifying a covariate-response curve
without explicitly acknowledging how this arises from habitat-specific
movement and demography (virtualspecies), or model movement using a
dispersal kernel (RangeShifter, STEPS) or a least-cost path algorithm
(SMS, STEPS). By contrast, `marlin` uses a continuous-time Markov chain
movement model, which integrates multiple paths (including their
path-dependent probability based on intervening habitat types and
species preferences) while using scale-free parameters that can be
measured experimentally in laboratory or tagging studies.

Simplified bio-economic models such as those used in @hastings2017,
@sala2021, and @cabral2019 can be applied at scale and provide
analytically tractable results, but as a result must abstract over many
bio-economic dynamics that can be important for more tactical
applications. `marlin` allows for more realistic representations of spatial social-ecological systems while maintaining processing speed.

In sum, `marlin` complements the existing spatial marine modeling
literature by allowing scientists, decision makers, and other
stakeholders to efficiently and productively examine the impacts of
realistic bio-economic dynamics on academic and applied problems. For example, scientists could use `marlin` to consider the relative importance of different biological variables in affecting policy outcomes. Managers could use `marlin` to assess the potential of proposed MPA networks to achieve stated objectives, such as maximizing conservation gains for a threatened species while minimizing or removing negative impacts to affected fishing fleets.
